\chapter{Background}
\label{chap:background}

\section{High-Frequency Trading and the Limit Order Book}

High-Frequency Trading (HFT) is a specialized form of algorithmic trading that 
executes orders at millisecond or microsecond timescales. 
HFT strategies are characterised by high order-to-trade ratios, exceedingly 
short holding periods, and a reliance on co-located servers running directly 
adjacent to exchange matching engines~\citep{aldridge2013hft}. 
Leading proprietary trading firms, such as Optiver and Citadel Securities, secure 
direct market data feeds to bypass the latency overhead of standard retail brokers. 
These firms leverage their ultra-low latency infrastructure to execute a variety 
of latency-sensitive strategies~\citep{menkveld2013hft}. 
This market-design and speed interaction is also synthesised at the overall
market-structure level by~\citet{ohara2015hfmicrostructure}.
This includes spread capture (acting as a market maker by buying at the bid and 
selling at the ask), avoiding adverse selection (cancelling stale quotes before 
they are traded against by faster participants), and statistical arbitrage 
(exploiting fleeting price discrepancies across multiple exchange venues).
As \citet{avellaneda2008hftlob} formalise, these strategies depend directly on 
how quickly the limit order book can be updated and acted upon. 
This is because even small delays can erode spread capture and increase 
adverse-selection risk. 

A core technological component of this ecosystem is the Limit Order Book (LOB), 
illustrated in Figure~\ref{fig:lob_diagram}. 
The LOB is a centralised electronic record of all active limit orders - instructions 
to buy or sell a specific quantity of an asset at a specified price or 
better~\citep{gould2013lob}. 
These orders are grouped by price levels and divided into two sides: the 
\hyperref[app:glossary]{\textit{bid}} (buyers) and the \hyperref[app:glossary]{\textit{ask}} or \textit{offer} (sellers). 
The highest price a buyer is willing to pay is the best bid, while the lowest 
price a seller will accept is the best ask. 
The numerical difference between these two values is known as the spread 
(visually separating the two sides in Figure~\ref{fig:lob_diagram}), which 
acts as a primary indicator of market liquidity and volatility. 
The midpoint between the best bid and best ask provides a theoretical ``fair value'' 
for the asset at any given microsecond. 
Strategic order submission and cancellation behaviour that drives spread and 
depth evolution is formalised by \citet{rosu2009dynamiclob}.

\input{figures/lob_diagram.tex}
\vspace{1em}

While the official limit order book is maintained by the exchange matching engine, 
trading firms are responsible for reconstructing their own internal "shadow" books. 
Exchanges broadcast market events via raw binary protocols like ITCH or FIX, and 
quant firms must rapidly parse these feeds to rebuild the current book state in 
their own local memory. 
The performance of an HFT algorithm depends entirely on how quickly this internal 
shadow book can process events. 
This makes the underlying data structures and software architecture 
paramount.

\section{Hardware Architecture}

To achieve true ultra-low latency, trading systems must be designed to cater 
directly to the physical realities of the underlying microprocessor. 
This approach is widely known as Mechanical Sympathy~\citep{fowler2011lmax}. 
The core philosophy dictates that understanding the hardware architecture of a 
machine enables software to operate in harmony with it, unlocking maximum 
potential throughput. 
Rather than treating the CPU as an abstract resource, HFT software treats hardware 
features - such as cache hierarchies and processor pipelines - as first-class 
design constraints.
\subsection{Cache Hierarchy and Memory Access}

The cache is the primary hardware component to optimise for when 
striving for ultra-low latency. 
It relies on the concepts of temporal and spatial locality to reduce memory operation times. 
Temporal locality dictates that recently accessed memory locations are highly 
likely to be accessed again in the near future. 
Spatial locality assumes that memory locations physically adjacent to a recently 
accessed address will also be needed soon. 
This behaviour closely aligns with the Pareto principle, where 80\% of memory 
accesses are typically concentrated within just 20\% of an application's data. 

To capitalise on this, the memory subsystem is organised into a rigid hierarchical 
structure, as visualised in Figure~\ref{fig:cache_hierarchy}. 
The \hyperref[app:glossary]{L1 cache} is the smallest storage layer but resides physically closest to the 
processor core, offering extremely fast access times of roughly 1 - 2 nanoseconds. 
Subsequent caches (L2 and L3) are progressively larger but sit further away, 
incurring proportionately longer fetch times. 
At the bottom of this hierarchy lies the main system memory (RAM). 
Fetching data from main memory is significantly slower relative to the cache, 
routinely taking over 100 nanoseconds and stalling the CPU pipeline entirely
~\citep{patterson2013computer}.

When the CPU fetches data from memory, it does not retrieve individual bytes; 
instead, it loads contiguous blocks known as \hyperref[app:glossary]{cache lines} (typically 64 bytes). 
This mechanism perfectly leverages spatial locality, ensuring that contiguous 
data structures - like flat arrays - can be sequentially processed with minimal misses. 
Conversely, non-contiguous allocations (such as pointer chasing in tree structures) 
break spatial locality, forcing the CPU to continuously stall while waiting for 
slow main memory fetches.

\input{figures/cache_hierarchy.tex}
\vspace{1em}
\subsection{Branch Prediction}
Modern CPUs process computation through a specific sequence of stages known as 
the instruction pipeline, conceptualised in Figure~\ref{fig:pipeline_flush}. 
While architectures vary, a classic pipeline consists of four fundamental stages: 
Fetch, Decode, Execute, and Write-back~\citep{patterson2013computer}. 
When the CPU encounters a conditional branch (an \texttt{if}-\texttt{else} statement) 
during the Fetch phase, it does not actually know which path to take until the 
Execute phase resolves the condition. 
At this decision point, the CPU has two choices: wait for the execution to finish, 
or guess the outcome. 
Waiting would stall the pipeline completely, wasting upward of 15 to 20 clock 
cycles while the CPU sits idle doing nothing. 

To avoid this massive waste of computational time, CPUs rely on a mechanism 
called a branch predictor. 
Instead of waiting, the branch predictor guesses which path the program will take 
and immediately begins speculatively fetching and decoding those instructions. 
If the guess is correct, execution continues seamlessly with zero wasted cycles. 
However, if the predictor guesses incorrectly, the CPU must flush all the 
speculative, half-finished work from the pipeline and restart fetching from the 
correct address. 
This pipeline flush incurs a severe penalty, typically costing modern architectures 
like the AMD Zen 4 between 15 and 20 cycles of completely lost time~\citep{fog_microarchitecture}. 

Modern predictors, such as the TAGE (TAgged GEometric) or perceptron-based 
predictors built into current AMD and Intel chips, are incredibly sophisticated 
and routinely achieve over 95\% accuracy~\citep{seznec2006case}. 
Mathematically, the advantage is clear: assuming a 15-cycle misprediction penalty, 
a 95\% accurate predictor yields an average penalty of just $0.75$ cycles per branch 
($0.95 \times 0 + 0.05 \times 15$), vastly outperforming a strategy of always 
stalling for 15 cycles. 

\input{figures/pipeline_flush.tex}
\vspace{1em}

However, despite this impressive average-case mathematical advantage, unpredictable 
branches are highly detrimental to ultra-low latency systems. 
In HFT, average performance is secondary to worst-case tail latency (the outermost 
response-time percentiles, such as the 99.9th percentile, which dictate the absolute 
maximum time a system takes to react under stress)~\citep{dean2013tail}. 
If complex business logic (such as an unpredictable conditional order fill) resides 
on the critical path, the inevitable 5\% of mispredictions will cause unpredictable, 
nanosecond-scale execution spikes. 
Removing unpredictable branches entirely is therefore a primary engineering goal 
to ensure deterministic, low-variance execution.

\section{Data Structures for Order Books}

The choice of underlying data structure fundamentally dictates the baseline 
performance of any limit order book. 
While theoretical algorithmic complexity provides a mathematical framework, the 
physical realities of modern hardware - namely the cache hierarchy and branch 
predictor - often subvert classical computer science assumptions.

\subsection{Tree-Based Implementations}
Historically, many order books have been implemented using tree-based structures 
such as Red-Black trees (often conceptually exposed as \texttt{std::map} in C++) 
to manage price levels. 
These structures offer a theoretical $\mathcal{O}(\log N)$ time complexity for 
insertions, deletions, and lookups. 
Their primary advantage is implementation simplicity, as a map intuitively models 
the domain of a limit order book by directly associating a price level (the key) 
with a queue of resident orders (the value). 
Furthermore, trees dynamically and gracefully handle massive, arbitrary price leaps 
without requiring predefined price boundaries. 
However, tree nodes are invariably allocated individually on the dynamic heap. 
As the order book processes thousands of intertwined updates, these nodes become 
fragmented across disparate locations in main memory. 
Consequently, traversing a tree to find a specific price level results in rapid, 
unpredictable pointer chasing. 
Each pointer dereference frequently triggers a cache miss, forcing the CPU to 
repeatedly stall while waiting for main memory fetches.

\subsection{Flat Array Structures}
To explicitly combat heap fragmentation, modern HFT architectures favour 
contiguous, flat structures such as pre-allocated arrays or 
vectors~\citep{bilokon2023cpp}. 
Although inserting an element into a sorted array incurs an $\mathcal{O}(N)$ 
worst-case complexity (due to shifting elements to maintain chronological order), 
this metric is theoretically inferior to a tree's $\mathcal{O}(\log N)$. 
However, because arrays are strictly contiguous in memory, traversing them allows 
the CPU's hardware prefetcher to eagerly load subsequent elements into the L1 
cache before they are even requested~\citep{patterson2013computer}. 
For the relatively small, localised working sets typical of an active limit order 
book, the sheer speed of contiguous localised memory scans vastly outweighs the 
algorithmic overhead. 
In practice, an $\mathcal{O}(N)$ cache-friendly array scan routinely outperforms 
an $\mathcal{O}(\log N)$ tree traversal for latency-critical workloads. 
The severe trade-off for this speed is architectural rigidity: to achieve true 
$\mathcal{O}(1)$ direct array indexing based on price, the book must strictly map 
discrete \textit{fixed-tick} price intervals to precise internal array offsets, 
sacrificing the dynamic flexibility of a tree.

\subsection{Memory Pooling}

The final structural consideration is how memory is allocated during live trading. 
Standard dynamic allocation functions like \texttt{std::malloc} or the \texttt{new} 
operator require the operating system to search for free heap space and occasionally 
request more memory pages from the kernel. 
This process requires \hyperref[app:glossary]{context switching} and is inherently non-deterministic. 
Executing a dynamic allocation on the critical trading path is therefore fatal to 
predictability because it randomly introduces severe microsecond-scale execution spikes. 

\input{figures/memory_pool.tex}

As illustrated in Figure~\ref{fig:memory_pool}, to bypass the operating system 
entirely during the trading day, high-performance systems use Memory Pools - 
large contiguous blocks of memory allocated entirely during system startup. 
Typically, this is achieved by instructing the OS to persistently map a massive, 
pre-sized memory region (commonly via the POSIX \texttt{mmap} system call) capable 
of securely holding millions of queued objects without exhausting system bounds~\citep{love2013linux}. 
At runtime, inserting a new order simply claims a pre-allocated object from the 
local pool via a fast pointer bump, guaranteeing deterministic, $\mathcal{O}(1)$ 
nanosecond execution. 

From a software engineering perspective, this profound architectural change is 
encapsulated behind standard interfaces. The memory pool acts as a drop-in 
replacement, exposing the same high-level API methods (e.g., \texttt{insert}, 
\texttt{erase}) as a traditional \texttt{std::map}, allowing the trading 
engine to remain agnostic to the underlying memory mechanics. 
This, however, shifts the burden entirely onto the developer, mandating manual 
capacity management and the implementation of custom logic to handle aggressive, 
unforeseen out-of-memory pool exhaustions.

\vspace{1em}

\section{\hyperref[app:glossary]{Lock-Free} Concurrent Queues}

\subsection{Exchange Architecture and Queue Modelling}
In financial markets, trading firms transmit orders to exchanges 
across high-speed optical networks.
Upon ingress, an exchange must safely and deterministically sequence 
these concurrent, intertwining arriving messages to enforce 
absolute chronological fairness. 
Physical exchanges achieve this sequencing via hardware - performing 
nanosecond packet timestamping directly on the ingress Network 
Interface Card (NIC) before funnelling messages through a 
hardware-level arbiter~\citep{harrisequity}. 
However, to stress-test the isolated performance algorithms 
of the matching engine without requiring proprietary networking hardware, 
our benchmark methodology strictly models this exchange boundary 
as a system of concurrent software queues. 
Network-facing threads act as producers that push validated 
messages into a gateway queue.
The matching engine acts as the sole consumer, 
polling the queue to execute orders sequentially.

\input{figures/network_funnel.tex}

\subsection{SPSC and MPMC Implementations}
To evaluate this hand-off boundary, 
the architecture must forbid the use of operating-system locks. 
For a 1-to-1 thread mapping (one network thread, one matching engine), 
a custom \hyperref[app:glossary]{Single-Producer Single-Consumer (SPSC)} ring buffer 
is the natural model. 
As illustrated in Figure~\ref{fig:spsc_queue}, an SPSC queue coordinates 
thread access using two entirely independent atomic variables: 
a \textit{Head} index pushed exclusively by the producer, 
and a \textit{Tail} index strictly advanced by the consumer. 

\input{figures/spsc_queue.tex}

To validate gateway scalability against concurrent trading clients, 
the benchmarking architecture must support Many-to-One messaging. 
For multi-client configurations, the system integrates 
\texttt{moodycamel::ConcurrentQueue}~\citep{moodycamel_concurrentqueue}. 
This \hyperref[app:glossary]{Multi-Producer Multi-Consumer (MPMC)} structure allows the 
evaluation suite to simulate gateway contention from multiple 
concurrent threads and observe how throughput scales with producer count.

\input{figures/mpmc_contention.tex}

\subsection{Atomic Operations and Cache Coherence}
Lock-free queues remove standard mutual exclusion locks (e.g., \texttt{std::mutex}) 
to avoid the risk of threads being put to sleep by the operating system. 
By eliminating context switching, they improve determinism and reduce latency 
spikes during high contention. Instead, they rely on atomic machine 
instructions, which allow a core to update a variable without 
relinquishing the CPU.

However, writing to an atomic variable requires a core to obtain exclusive 
ownership of the memory location. This triggers the hardware's 
\hyperref[app:glossary]{\textit{MESI} protocol}~\citep{patterson2013computer}, which forces all other 
cores to invalidate their local cache copies. 
Communication across the CPU bus to synchronise these states is 
orders of magnitude slower than a local L1 cache access.

Performance is further degraded by \hyperref[app:glossary]{\textit{false sharing}}. This occurs when 
unrelated variables—such as the producer's \textit{Head} index and the 
consumer's \textit{Tail} index—accidentally occupy the same 64-byte 
hardware cache line. Because the CPU manages memory in these 64-byte blocks, 
cores will constantly invalidate each other's cache even though they 
are updating different indices.

To prevent this, high-performance implementations use padding and 
C++20 alignment specifiers to ensure each index resides on its own 
unique cache line, as demonstrated in the following snippet:

\begin{lstlisting}[language=C++, basicstyle=\small\ttfamily, frame=single, keywordstyle=\color{blue}]
struct SPSCQueue {
    // Force separation into different 64-byte cache lines
    alignas(64) std::atomic<size_t> head; // Pushed by Producer
    alignas(64) std::atomic<size_t> tail; // Advanced by Consumer
};
\end{lstlisting}

The spatial management of queue indexes is a prerequisite for scaling 
lock-free performance on multi-core hardware. 
These techniques are grounded in foundational non-blocking algorithm theory, 
notably the single-producer single-consumer and multi-producer queues 
formulated by~\citet{michael1996simple}, which define the safety properties 
required for wait-free ingestion.


The trading industry employs a diverse array of communication protocols, 
optimised for specific trade-offs between flexibility and performance. 
These range from flexible, text-based encodings to rigid, dense 
binary formats. In all cases, these protocols prioritise data integrity, 
incorporating mandatory checksums and sequence numbers to 
ensure orders are recovered and reconciled during 
network instability~\citep{fix42_spec}.
This protocol-level focus is directly relevant in low-latency markets, where
high rates of order submission and cancellation increase message volume. In these cases
latency from network receive, protocol parsing, and queue waiting becomes a
significant part of end-to-end execution time~\citep{hasbrouck2013lowlatency}.

In our exchange implementation, we use the Financial Information eXchange 
(FIX) \hyperref[app:glossary]{protocol}, the global standard for order routing. 
FIX is a session-based protocol communicated over TCP/IP 
to ensure reliable, ordered delivery of market messages. 
It employs a tag-value encoding scheme where each field is represented 
by a unique numeric tag, an equals sign, and the field's value, as 
illustrated in Figure~\ref{fig:fix_message_structure}.

\input{figures/fix_structure.tex}

As shown in Figure~\ref{fig:fix_message_structure}, each field is delimited 
by the \texttt{SOH} (Start of Header, ASCII \texttt{0x01}) 
character. 
The protocol includes a mandatory \textit{Standard Header} containing 
the message type (Tag 35), a \textit{Message Body} containing order 
details (e.g., Symbol, Price, Quantity), and a \textit{Standard Trailer} 
containing a modulo-256 checksum (Tag 10). 
While the tag-value format is highly extensible and human-readable, 
parsing these strings necessitates repeated branching and integer 
conversions, making it significantly slower than modern binary 
encodings like Simple Binary Encoding (SBE)~\citep{dean2013tail}.

\subsection{The Standard Linux Network Stack}
When a host receives a FIX message encapsulated within 
a TCP/IP packet, it must traverse the standard Linux kernel networking 
stack. This journey, visualised in Figure~\ref{fig:kernel_network_pipeline}, 
is a rigid, multi-stage pipeline designed for throughput and safety rather 
than deterministic latency.

\input{figures/kernel_pipeline.tex}

Upon arrival at the Network Interface Card (NIC), the packet is 
written via Direct Memory Access (DMA) to a ring buffer in system memory. 
The NIC then triggers a hardware interrupt (Hard IRQ), forcing the 
CPU to pause its current task and branch to the kernel's interrupt 
handler. 
The kernel then switches to a Software Interrupt (SoftIRQ) 
context via the NAPI (New API) polling mechanism to drain the ring 
buffer, as detailed by~\citet{love_kernel}. 
The kernel allocates an \texttt{sk\_buff} data structure 
to wrap the packet, performs housekeeping and routing (IP layer), 
and eventually appends the data to the socket's receive queue.

The User Application must issue a system call, such as 
\texttt{recvfrom}, to retrieve the data. This triggers an 
expensive \textit{context switch} from User Space to Kernel Space, 
where the kernel copies the data from its internal buffer to the 
application memory. 
Because the kernel is responsible for scheduling thousands of unrelated 
tasks across the CPU, a packet may sit in the socket queue for 
unpredictable durations if the kernel is busy with book-keeping or 
other interrupts. 
This accumulation of interrupts, memory allocations, and context 
switches is the fundamental source of the microsecond-scale 
\textit{jitter} (latency variance) that plagues standard network 
sockets in HFT.

\subsection{Kernel-Bypass Networking}
To mitigate the systemic non-determinism of the Linux kernel, 
trading firms employ kernel-bypass technologies. 
Kernel-bypass is an architectural shift that allows the user-space 
application to talk directly to the NIC hardware, circumventing 
the OS networking stack, as compared in 
Figure~\ref{fig:kernel_bypass_comparison}.

\input{figures/kernel_bypass.tex}

By using specialised libraries such as the Data Plane Development 
Kit (DPDK) or Solarflare's OpenOnload~\citep{redhat2012_openonload}, 
the application takes ownership of the NIC's hardware 
resources. 
Instead of waiting for interrupts, a dedicated CPU core is 
frequently assigned to "spin" (poll) the NIC's memory-mapped 
ring buffer in a continuous loop. 
This achieves \hyperref[app:glossary]{\textit{Zero-Copy}} delivery, where packets are read 
directly into application memory without kernel intervention or 
system calls. 
This removal of the OS bottleneck ensures network latency is 
consistent and deterministic, an essential prerequisite for 
validating the performance differences of 
order-book data structures which would otherwise be masked by 
kernel jitter.

\section{Related Work}
Prior literature can be grouped into market-microstructure theory,
systems-design practice, and performance-engineering methodology, each with
different strengths and blind spots.

Market-microstructure sources establish why order books matter economically.
\citet{harrisequity}, \citet{aldridge2013hft}, and \citet{menkveld2013hft}
explain liquidity formation, adverse selection, and the strategic role of fast
execution. \citet{gould2013lob} provides a broad empirical and modelling survey
of limit-order-book dynamics. However, these works are not primarily designed as
implementation-level benchmarking studies, so they offer limited guidance on how
specific in-memory data-structure choices affect nanosecond-scale matching-path
latency.
Recent exchange-scale evidence by \citet{aquilina2022armsrace} further shows that
small latency advantages can have economically significant effects.

Systems-oriented sources describe low-latency architectural patterns but are
less controlled from an academic benchmarking perspective. The LMAX work
\citep{fowler2011lmax,thompson2011disruptor} and more recent systematic
evaluations by \citet{bilokon2023cpp} demonstrate the practical value
of single-writer flows, cache-aware patterns, and the "boring" (deterministic) 
latency profiles required for HFT. Library-level
artefacts such as moodycamel \citep{moodycamel_concurrentqueue} show strong
queue throughput under synthetic conditions, but they do not by themselves
establish end-to-end exchange latency behaviour across heterogeneous order-book
implementations.

Performance-methodology references provide the measurement principles needed to
connect these domains. \citet{gregg2013systems} and \citet{dean2013tail}
emphasise observer-effect control, percentile-aware reporting, and bottleneck
localisation rather than relying on single aggregate means. This is directly
relevant to HFT benchmarking, where tail behaviour and queueing effects can
change engineering decisions even when average latency appears similar.

The main gap is therefore not a lack of discussion on HFT speed, but a lack
of reproducible, decomposition-driven comparisons that evaluate both
algorithmic and end-to-end effects in one framework. Existing literature either
focuses on financial behaviour without implementation granularity, or on
low-latency engineering patterns without controlled cross-structure comparison
in a unified exchange-style pipeline. This dissertation addresses that gap by
benchmarking five order-book designs under semantically equivalent workloads in
both direct mode and gateway mode, then decomposing network, queue, and engine
latency to identify where masking effects occur.
