\documentclass[12pt,a4paper]{report}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage[numbers]{natbib}
\usepackage{setspace}
\geometry{margin=1in}
\onehalfspacing

\title{High-Frequency Trading Order Book Benchmarking in C++20}
\author{Your Name}
\date{\today}

\begin{document}

\pagenumbering{roman}
\maketitle
\tableofcontents
\listoffigures
\listoftables

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
A C++20 exchange infrastructure is presented for benchmarking five order book data structure strategies across direct and gateway operating modes. The framework contains a TCP order gateway, an incremental FIX parser, a lock-free single-producer single-consumer queue, and a single-threaded matching engine connected to five interchangeable order book implementations. Results show that direct-indexed array access achieves the strongest direct-mode latency under most scenarios, with representative mean latency near 174 ns in mixed conditions, while sorted-vector insertion degrades sharply in dense books, reaching approximately 13,748 ns in dense-full workloads. End-to-end gateway measurements remain dominated by network and transport overhead, with loopback TCP path latency on the order of 4.5 ms, reducing the practical contribution of order-book-internal differences to total latency. Findings support two conclusions. First, cache locality and elimination of pointer indirection dominate low-latency matching performance when algorithmic work is isolated. Second, system-level bottlenecks can mask data structure gains unless transport and parsing overhead are reduced. The project contributes a reproducible benchmark harness, scenario-driven workload generation, and decomposed Net/Que/Eng timing suitable for future experiments on kernel-bypass networking and multi-producer ingestion.

\clearpage
\pagenumbering{arabic}

\include{chapters/01_introduction}
\include{chapters/02_background}
\include{chapters/03_system_design}
\include{chapters/04_implementation}
\include{chapters/05_testing_evaluation}
\include{chapters/06_conclusions}

\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\chapter{LockFreeQueue Implementation}
Add the full lock-free queue source listing here.

\chapter{ArrayOrderBook addOrder}
Add focused listing for index mapping and best-level cache updates.

\chapter{Matching Engine Timestamp Decomposition}
Add process order listing with send/receive/engine timestamps.

\chapter{FIX Incremental Parser}
Add parser loop showing partial-message buffering and checksum boundary detection.

\end{document}
