\documentclass[12pt,a4paper]{report}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage[numbers]{natbib}
\usepackage{setspace}
\geometry{margin=1in}
\onehalfspacing

\title{High-Frequency Trading Order Book Benchmarking in C++20}
\author{Your Name}
\date{\today}

\begin{document}

\pagenumbering{roman}
\maketitle
\tableofcontents
\listoffigures
\listoftables

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
A C++20 exchange infrastructure is presented for benchmarking five order book
data structure strategies across direct and gateway operating modes. The
framework contains a TCP order gateway, an incremental FIX parser, a lock-free
single-producer single-consumer queue, and a single-threaded matching engine
connected to interchangeable order book implementations. On the canonical
pinned-run dataset (2026-03-26), direct-indexed array access is the strongest
direct-mode implementation in six of seven scenarios, with mixed-scenario mean
latency of 210.72 ns and dense-full mean latency of 213.44 ns. In contrast,
sorted-vector insertion degrades sharply in dense books, reaching 16082.10 ns
in dense-full direct mode (75.3x slower than array). Gateway-mode latency is
dominated by network and transport overhead for non-vector rows (median Net
share 98.7\%), which reduces the visible contribution of internal order-book
differences to end-to-end timing. The evidence supports two bounded
conclusions. First, data-structure execution path and memory-access behavior
are the dominant factors when algorithmic work is isolated. Second, in
end-to-end gateway operation, system-level bottlenecks can mask data-structure
gains unless transport and parsing overhead are reduced. The project
contributes a reproducible benchmark harness, scenario-driven workload
generation, and decomposed Net/Que/Eng timing suitable for future experiments
on kernel-bypass networking and multi-producer ingestion.

\clearpage
\pagenumbering{arabic}

\include{chapters/01_introduction}
\include{chapters/02_background}
\include{chapters/03_system_design}
\include{chapters/04_implementation}
\include{chapters/05_testing_evaluation}
\include{chapters/06_conclusions}

\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\chapter{LockFreeQueue Implementation}
This appendix contains the lock-free queue implementation used in the
benchmark pipeline. The listing is provided in the submitted artefact bundle as
source code under the corresponding implementation directories.

\chapter{ArrayOrderBook addOrder}
This appendix points to the focused ArrayOrderBook insertion path, including
index mapping and best-level cache updates, as implemented in the submitted
source artefacts.

\chapter{Matching Engine Timestamp Decomposition}
This appendix references the matching-engine processing path and timestamp
decomposition used for Net/Que/Eng breakdown in the evaluation chapter.

\chapter{FIX Incremental Parser}
This appendix references the incremental FIX parser loop used for
partial-message buffering and checksum boundary detection in the gateway path.

\end{document}
